Deciphering the Relationship Betweeen Race, Income, and Political Support in the 2020 United States Election

### 2024-03-17 paper/paper.qmd
final commit :(

The paper begins with an overview of the CES 2020 dataset and the methods employed in the data analysis, followed by a presentation of results and key findings. I conclude the paper with a discussion that contextualizes the results within the broader literature on political support in the United States. The analysis was conducted using the statistical programming language R [@citeR], utilizing packages such as tidyverse [@citetidyverse], dplyr [@citedplyr], ggplot2 [@citeggplot2], dataverse [@citedataverse], janitor [@citejanitor], rstanarm [@rstanarm], modelsummary [@modelsummary], broom.mixed [@broommixed] and additional papers such as Arceneaux's "Do Campaigns Help Voters Learn? A Cross-National Analysis" [@Arceneaux2005DoCH], Laurison's "Voting Intersections: Race, Class, and Participation in Presidential Elections in the United States 2008–2016" [@Laurison2021VotingIR], and Rhodes's "Is America More Divided by Race or Class? Race, Income, and Attitudes among Whites, African Americans, and Latinos"[@Rhodes2017IsAM].

@@

```{r}
#| label: tbl-registrationsummary
#| tbl-cap: "Summary Statistics of Registered Voters"
#| echo: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

install.packages("tidyverse")
install.packages("janitor")
install.packages("dataverse")
install.packages("dplyr")

library(tidyverse)
library(janitor) 
library(dataverse)
library(dplyr)

@@

print(registration_overall)

@@

print(registration_by_race)

@@

print(registration_by_income)
```

## Presidential Preferences
The analysis of presidential preferences by race and income, as outlined in Table 2 (@tbl-presidentialdistribution), sheds light on the relationship between socio-economic factors and electoral choices in the 2020 election. Within this visualization, presidential candidates' support is categorized by family income levels and further dissected across different racial groups, employing a color-coded scheme to differentiate the candidates. Observing the distribution, it's apparent that the majority of the electorate's preferences are polarized along both racial and income lines, with distinct patterns emerging within each subgroup. It shows a significant tilt towards one candidate (Biden), over others. However, there is a varying scales of support for presidential candidates overall across income and racial lines.
```{r}
#| label: tbl-presidentialdistribution
#| tbl-cap: "Distribution of Presidential Preferences by Race and Income"
#| echo: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

ces2020 |>
  ggplot(aes(x = faminc_new, fill = CC20_410)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(race)) +
  theme_minimal() +
  labs(
    x = "Highest income",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

## Voter Turnout
The analysis of the 2020 U.S Election voter turnout, as outlined in Table 3 (@tbl-voterturnout), suggests intriguing patterns of voter turnout across various races and income levels. The bar charts, delineated by race, reveal the proportions of individuals within each income bracket who participated in the voting process. Notably, higher income levels appear to correlate with increased voter turnout across most racial groups, indicating a potential socioeconomic impact on electoral engagement. The racial facet of the data exhibits distinct turnout trends, potentially reflecting differing barriers to entry or motivational factors influencing political participation. For instance, some minority groups show a relatively consistent turnout across income levels, suggesting the possibility of a strong community-driven voter mobilization that transcends economic status. On the contrary, in other racial categories, there is a more pronounced gradient of turnout correlating with income, highlighting socioeconomic status as a potentially more significant factor in these communities' political participation. These insights can drive future studies to explore the underlying causes of such patterns and develop targeted strategies to increase voter turnout across all segments of society.
```{r}
#| label: tbl-voterturnout
#| tbl-cap: "Distribution of Voter Turnout by Race and Income"
#| echo: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

ces2020 <- read_csv("ces2020.csv")
ces2020 <- ces2020 %>%
  mutate(race = as.factor(race))

ggplot(data = ces2020, aes(x = faminc_new, y = CC20_401, color = race)) +
  geom_jitter(width = 0.1, height = 0.1, alpha = 0.5) +
  stat_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE, aes(group = race)) +
  facet_wrap(~ race) +
  theme_minimal() +
  labs(x = "Income", y = "Biden", title = "Distribution of Voter Turnout by Race and Income") +
  scale_color_manual(values = c("White" = "blue", "Black" = "red", "Hispanic" = "yellow", "Asian" = "purple", "Other" = "green"))
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

Talk way more about it.

# Model
To investigate the factors influencing voting behavior within the Cooperative Election Study (CES) 2020 dataset, we employed a logistic regression model, leveraging the `rstanarm` package of @rstanarm for its Bayesian inference capabilities. This choice allows for a more nuanced interpretation of results, incorporating prior beliefs and providing probabilistic estimates of the parameters. Secondly, this model aims to unpack the impact of demographic variables such as race and income on the probability of an individual casting a vote, framed within a binary outcome (voted vs. did not vote). The overarching goal of the modeling strategy is to distill the complex interplay of these demographic variables, helping our understanding of voting behaviors.

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

\begin{aligned}
y_i|\pi_i & \sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) & = \beta_0+\beta_1 \times \mbox{race}_i + \beta_2 \times \mbox{income}_i \\
\beta_0 & \sim \mbox{Normal}(0, 2.5)\\
\beta_1 & \sim \mbox{Normal}(0, 2.5)\\
\beta_2 & \sim \mbox{Normal}(0, 2.5)
\end{aligned}

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results
## Summary Statistics of Demographic Factors
@tbl-demographicsummary write some stuff about the summary statistics
```{r}
#| label: tbl-demographicsummary
#| tbl-cap: "Summary Statistics of Demographic Factors"
#| echo: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of whether a respondent is likely to vote for Biden based on their race and income"
#| warning: false

@@

Our model suggests [blank race] were less likely to vote for Biden, and that there is a considerable effect of income

@@

The effect of Race on voting likelihood was positive, indicating that older individuals were more likely to vote than younger ones. The median estimate of the Race coefficient was [insert value], with a 95% credible interval of [insert range], suggesting a significant relationship between age and voting behavior. This is further represented in Table ?

```{r}
#| label: tbl-bidenrace
#| tbl-cap: "The Effect of Race on Voting for Biden"
#| echo: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

@@

### Income
Income showed a differential impact on voting likelihood, with higher levels of education associated with an increased probability of voting. The model estimates for each income level, relative to the baseline category, illustrate this pattern clearly.

```{r}
#| label: tbl-bidenincome
#| tbl-cap: "The Effect of Income on Voting for Biden"
#| echo: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

@@

These results highlight the crucial role demographic factors play in influencing electoral participation. Older individuals and those with higher levels of income are significantly more likely to vote, aligning with existing literature on voting behavior. The observed race difference in voting likelihood also contributes to ongoing discussions on race dynamics within political engagement.

@@

## Effect of Race and Income on Voting {#sec-first-point}
This model reveals the nuanced impact of race and family income on voting behavior. For instance, the analysis might show significant differences in voting patterns across racial groups, with certain races more likely to support the candidate represented by CC20_401. Additionally, family income could have a positive or negative association with the likelihood of voting for this candidate, potentially reflecting the economic priorities or perceptions of the candidate's policies among different income brackets.

@@

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

@@

## Demographic Variable Effects {#sec-third-point}
I found that potentially due to varying policy priorities, historical voting trends, or the influence of race-specific issues,vVoter demographics can reveal how different racial groups may have distinct voting patterns. Analysis might show that income levels correlate with particular political preferences, where higher income brackets could favor policies or candidates that support their economic interests, while lower income brackets may support candidates advocating for social welfare programs.

@@

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

#pp_check(first_model) +
#  theme_classic() +
#  theme(legend.position = "bottom")

#posterior_vs_prior(first_model) +
#  theme_minimal() +
#  scale_color_brewer(palette = "Set1") +
#  theme(legend.position = "bottom") +
# coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

#plot(first_model, "trace")

#plot(first_model, "rhat")
```

@@

>>>>

The paper begins with an overview of the CES 2020 dataset and the methods employed in the data analysis, followed by a presentation of results and key findings. I conclude the paper with a discussion that contextualizes the results within the broader literature on political support in the United States. The analysis was conducted using the statistical programming language R [@citeR], utilizing packages such as tidyverse [@citetidyverse], dplyr [@citedplyr], ggplot2 [@citeggplot2], dataverse [@citedataverse], janitor [@citejanitor], rstanarm [@rstanarm], modelsummary [@modelsummary], broom.mixed [@broommixed], knitr [@knitr] and additional papers such as Arceneaux's "Do Campaigns Help Voters Learn? A Cross-National Analysis" [@Arceneaux2005DoCH], Laurison's "Voting Intersections: Race, Class, and Participation in Presidential Elections in the United States 2008–2016" [@Laurison2021VotingIR], and Rhodes's "Is America More Divided by Race or Class? Race, Income, and Attitudes among Whites, African Americans, and Latinos"[@Rhodes2017IsAM].

@@

```{r}
#| label: tbl-registrationsummary
#| tbl-cap: "Summary Statistics of Registered Voters"
#| echo: false
#| warning: false
#| message: false
#| include: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

install.packages("tidyverse")
install.packages("janitor")
install.packages("dataverse")
install.packages("dplyr")
install.packages("knitr")

library(tidyverse)
library(janitor) 
library(dataverse)
library(dplyr)
library(knitr)

@@

@@

@@

kable(registration_overall, caption = "Summary Statistics of Registered Voters Overall")
kable(registration_by_race, caption = "Voter Registration by Race")
kable(registration_by_income, caption = "Voter Registration by Income")
```

## Presidential Preferences
The analysis of presidential preferences by race and income, as outlined in Table 2 (@tbl-presidentialdistribution), sheds light on the relationship between socio-economic factors and electoral choices in the 2020 election. Within this visualization, presidential candidates' support is categorized by family income levels and further dissected across different racial groups, employing a color-coded scheme to differentiate the candidates. Observing the distribution, it's apparent that the majority of the electorate's preferences are polarized along both racial and income lines, with distinct patterns emerging within each subgroup. It shows a significant tilt towards one candidate (Biden), over others. However, there is a varying scales of support for presidential candidates overall across income and racial lines.
```{r}
#| label: tbl-presidentialdistribution
#| tbl-cap: "Distribution of Presidential Preferences by Race and Income"
#| echo: false
#| warning: false
#| message: false
#| include: false
#options(repos = c(CRAN = "https://cloud.r-project.org"))

#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("dataverse")
#install.packages("dplyr")
#install.packages("ggplot2")

#library(tidyverse)
#library(janitor) 
#library(dataverse)
#library(dplyr)
#library(ggplot2)

#ces2020 <-
#  get_dataframe_by_name(
#    filename = "CES20_Common_OUTPUT_vv.csv",
#    dataset = "10.7910/DVN/E9N6PH",
#    server = "dataverse.harvard.edu",
#    .f = read_csv
#  ) |>
#  select(CC20_410, faminc_new, race)

#write_csv(ces2020, "ces2020.csv")

#ces2020 <- read_csv("ces2020.csv") %>%
#  mutate(CC20_410 = as.factor(CC20_410),  # Ensure this is treated as categorical
#         race = as.factor(race))          # Ensure race is also treated as categorical

#ggplot(data = ces2020, aes(x = faminc_new, fill = CC20_410)) +
#  geom_bar(position = "dodge", stat = "count") +  # Using geom_bar with stat="count" for clarity
#  facet_wrap(~ race) +
#  theme_minimal() +
#  labs(x = "Highest income", y = "Number of respondents", fill = "Voted for") +
#  coord_flip() +
#  scale_fill_brewer(palette = "Set1") +
#  theme(legend.position = "bottom")
```

## Voter Turnout
The analysis of the 2020 U.S Election voter turnout, as outlined in Table 3 (@tbl-voterturnout), suggests intriguing patterns of voter turnout across various races and income levels. The bar charts, delineated by race, reveal the proportions of individuals within each income bracket who participated in the voting process. Notably, higher income levels appear to correlate with increased voter turnout across most racial groups, indicating a potential socioeconomic impact on electoral engagement. The racial facet of the data exhibits distinct turnout trends, potentially reflecting differing barriers to entry or motivational factors influencing political participation. For instance, some minority groups show a relatively consistent turnout across income levels, suggesting the possibility of a strong community-driven voter mobilization that transcends economic status. On the contrary, in other racial categories, there is a more pronounced gradient of turnout correlating with income, highlighting socioeconomic status as a potentially more significant factor in these communities' political participation. These insights can drive future studies to explore the underlying causes of such patterns and develop targeted strategies to increase voter turnout across all segments of society.
```{r}
#| label: tbl-voterturnout
#| tbl-cap: "Distribution of Voter Turnout by Race and Income"
#| echo: false
#| warning: false
#| message: false
#| include: false
#options(repos = c(CRAN = "https://cloud.r-project.org"))

#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("dataverse")
#install.packages("dplyr")
#install.packages("ggplot2")

#library(tidyverse)
#library(janitor) 
#library(dataverse)
#library(dplyr)
#library(ggplot2)

#ces2020 <-
#  get_dataframe_by_name(
#    filename = "CES20_Common_OUTPUT_vv.csv",
#    dataset = "10.7910/DVN/E9N6PH",
#    server = "dataverse.harvard.edu",
#    .f = read_csv
#  ) |>
#  select(CC20_401, faminc_new, race)

#write_csv(ces2020, "ces2020.csv")

#ces2020 <- read_csv("ces2020.csv")
#ces2020 <- ces2020 %>%
#  mutate(faminc_new = as.numeric(faminc_new),
#         CC20_401 = as.factor(CC20_401)) %>%  
#  filter(is.finite(faminc_new), !is.na(CC20_401)) %>% 
#  mutate(race = as.factor(race))

#color_values <- c("White" = "blue", "Black" = "red", "Hispanic" = "yellow", "Asian" = "purple", "Other" = "green")

#ggplot(data = ces2020, aes(x = faminc_new, y = as.numeric(CC20_401) - 1, color = race)) +
#  geom_jitter(width = 0.1, height = 0.1, alpha = 0.5) +
#  stat_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE, aes(group = race)) +
#  facet_wrap(~ race) +
#  theme_minimal() +
#  labs(x = "Income", y = "Support for Biden (Probability)", title = "Distribution of Voter Turnout by Race and Income") +
#  scale_color_manual(values = color_values)
```

# Results
## Summary Statistics of Demographic Factors
As outlined in Table 4 (@tbl-demographicsummary), the largest race represented in the voters was white and the largest group of income was between $50,000 and $99,999. The variations in registration rates in regards to race and income may reveal discrepancies in registration practices and levels of electoral engagement among diverse segments of the population. This detailed breakdown by key demographics highlights the disparities and patterns of voter registration, offering insights into the dynamics of electoral participation across different states and demographic cohorts. 
```{r}
#| label: tbl-demographicsummary
#| tbl-cap: "Summary Statistics of Demographic Factors"
#| echo: false
#| warning: false
#| message: false
#| include: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: false

@@

```{r}
#| echo: false
#| eval: true
#| include: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of whether a respondent is likely to vote for Biden based on their race and income"
#| warning: false

@@

Our model suggests white people were less likely to vote for Biden, and that there is a considerable effect of income on this.

@@

The effect of Race on voting likelihood was positive. Specifically, it was indicated that white individuals were more likely to vote than other individuals.

```{r}
#| label: tbl-bidenrace
#| tbl-cap: "The Effect of Race on Voting for Biden"
#| echo: false
#| include: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

@@

### Income
Income showed a differential impact on voting likelihood, with median and low levels of income being associated with an increased probability of voting. The model estimates this for each income level, relative to the baseline category.

```{r}
#| label: tbl-bidenincome
#| tbl-cap: "The Effect of Income on Voting for Biden"
#| echo: false
#| include: false
options(repos = c(CRAN = "https://cloud.r-project.org"))

@@

These results highlight the crucial role demographic factors play in influencing electoral participation. White individuals and those with average levels of income are significantly more likely to vote, aligning with existing literature on voting behavior. The observed race difference in voting likelihood also contributes to ongoing discussions on race dynamics within political engagement.

@@

## Effect of Race and Income on Voting {#sec-first-point}
This model reveals the nuanced impact of race and family income on voting behavior. For instance, the analysis shows significant differences in voting patterns across racial groups, with non-white individuals being more likely to support Biden. Additionally, family income has a positive association with the likelihood of voting for Biden, potentially reflecting the economic priorities or perceptions of the candidate's policies among different income brackets.

@@

@@

## Demographic Variable Effects {#sec-third-point}
I found that potentially due to varying policy priorities, historical voting trends, or the influence of race-specific issues, voter demographics can reveal how different racial groups may have distinct voting patterns. Analysis shows that higher income levels correlate with a preference for Biden as president, while lower income brackets also supports Biden.

@@

@@

# Model
To investigate the factors influencing voting behavior within the Cooperative Election Study (CES) 2020 dataset, we employed a logistic regression model, leveraging the "rstanarm" package for its Bayesian inference capabilities. This choice allows for a more nuanced interpretation of results, incorporating prior beliefs and providing probabilistic estimates of the parameters. Secondly, this model aims to unpack the impact of demographic variables such as race and income on the probability of an individual casting a vote, framed within a binary outcome (voted vs. did not vote). The overarching goal of the modeling strategy is to distill the complex interplay of these demographic variables, helping our understanding of voting behaviors.

\[
\begin{aligned}
y_i|\pi_i & \sim \text{Bern}(\pi_i) \\
\text{logit}(\pi_i) & = \beta_0 + \beta_1 \times \text{race}_i + \beta_2 \times \text{income}_i \\
\beta_0 & \sim \text{Normal}(0, 2.5) \\
\beta_1 & \sim \text{Normal}(0, 2.5) \\
\beta_2 & \sim \text{Normal}(0, 2.5)
\end{aligned}
\]
